Computational methods in landscape ecology

### 2024-10-16 Sections/07-genetics.qmd
Updates review

Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs\; @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA\; @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

>>>>

Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [@Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [@Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [@Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
Analytical frameworks such as the *FOLDS* model (gene flow, genetic offsets, genetic load, dispersal and SDMs) [@Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis (RDA) [@Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

### 2024-10-16 Sections/08-spatial-planning.qmd
Updates review

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link, and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning, or through using simulation model outputs as input features [e.g., @Daigle2020OperationalizingEcological].

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example, by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similarly, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
<!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
One is the use of graph theory in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization] to guide conservation and restoration efforts.
These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem. 
However, due to software licenses and related cost, the optimal solver might not be always accessible.

>>>>

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity, and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example, by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similarly, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
<!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
One is the use of graph theory in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization] to guide conservation and restoration efforts.
These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation] such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem. 
However, due to software licenses and related cost the optimal solver might not be always accessible.

### 2024-10-16 Sections/09-conclusion.qmd
Updates review

Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].

>>>>

Since it emerged in the 1980s [@With2019EssentialsLandscape] the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

@@

Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for, e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].
